Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-05 Jackson_Liu_Temlock_Zhang_Lab2_Draft_Report.Rmd
Updated introduction section and EDA to align with new transformation of percentage variables
model 2 update

df <-df %>%
  mutate(    
    senior_per_100k = 100000*senior_pct,
    poverty_per_100k = 100000*poverty_pct, 
    unemployed_per_100k = 100000*unemployed_pct,
  )
```

@@

```{r, warning=FALSE, message=FALSE}
summary(df$senior_per_100k)

ggplot(data = df,
  mapping = aes(x= senior_per_100k)) +
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(10000,21000,1000)) +
  labs(title = 'Histogram of Seniors Distribution', 
  x = 'Number of seniors per 100,000 people', y = 'Count') + 
  scale_x_continuous(breaks=seq(10000,22000,1000))
```

@@

For model 2, state demographic information regarding racial composition(white percentage of total population /Black Percentage of Total Population), economic status(unemployed percentage/poverty percentage), and population seniors percentage was investigated to explore its potential connection to the COVID pandemic status.

#### 2-a) White Percentage of Total Population/ Black Percentage of Total Population

Variable of black percentage is stored as string variables. Moreover, three entries of black percentage of total population is "<0.01", which is to be dropped or replaced by values determined by extra resources for further analysis.

Here, the value of black percentage is replaced by 0 when it is "<0.01". Afterwards, the correlation of white percentage and black percentage is -0.42, which shows these two variables are correlated. Considering that the value of other race groups is much smaller, it is proper to only include white percentage in the regression model to explore the relation of care rate and racial composition.

The distribution of the white % is skewed, not an ideal normal distribution. While the distribution of the square of white % is more close to a normal distribution. In practice, this square could reflect the chance that the interaction happens between two white people.

```{r warning=FALSE}
black_num <- as.numeric(df$black_population_pct)
black_num[is.na(black_num)]<- 0
cor(df$white_population_pct, black_num)
```

```{r}
summary(df$white_population_pct)
```

```{r}
ggplot(data = df,
  mapping = aes(x= white_population_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(0,1,0.05)) +
  labs(title =  "Histogram of White Population Percentage", 
       x = "White percentage of total population", y = 'Count')+
  scale_x_continuous(breaks=seq(0, 1, 0.1))
```
```{r}
ggplot(data = df,
  mapping = aes(x= (white_population_pct)^2))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(0,1,0.05)) +
  labs(title =  "Histogram of Square of White Percentage", 
       x = "White percentage of total population^2", y = 'Count')+
  scale_x_continuous(breaks=seq(0, 1, 0.1))
```

#### 2-b) Percentage living under the federal poverty line (2018)

Correlation of pecentage living under poverty line and white percentage of total population shows that they are not significantly related. The distribution of the poverty percentage is not heavily skewed.

```{r}
summary(df$poverty_pct)
cor(df$poverty_pct, df$white_population_pct)
```

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6,21,1)) +
  labs(title =  "Histogram of Poverty Percentage", 
       x = "Percentage living under poverty line (2018)", y = 'Count')+
  scale_x_continuous(breaks=seq(6, 21, 1))
```

#### 2-c) Unemployed Percentage (2018)

Although, the distribution is of high peak in the middle of the range. Overall, it is not highly skewed or heavily tailed. In addition, the unemployed rate is correlated to the variable to the poverty rate, which is in an agreement with intuitive expectation.

```{r}
summary(df$unemployed_pct)
cor(df$unemployed_pct, df$poverty_pct)
```
```{r}
ggplot(data = df,
  mapping = aes(x= unemployed_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(0,10,1)) +
  labs(title =  "Histogram of Unemployed Percentage", 
       x = "Unemployed percentage (2018)", y = 'Count')+
  scale_x_continuous(breaks=seq(0, 10, 1))
```

#### 2-d) Seniors Percentage (Age 65+)

The seniors percentage (age 65+) is of a nearly normal distribution.

```{r}
summary(df$senior_pct)
```

```{r}
ggplot(data = df,
  mapping = aes(x= senior_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(0.10,0.22,0.01)) +
  labs(title =  "Histogram of Seniors Percentage (65+)", x = "Seniors percentage", 
       y = 'Count')+
  scale_x_continuous(breaks=seq(0.1, 0.25, 0.05))
```

### Model 2 Plots

No obvious trend could be easily observed or concluded from the plots regarding the dependent variables and candidate covariates.

```{r message=FALSE}
df %>% 
  ggplot(aes(white_population_pct, case_rate_100k, color = log10(population_density))) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate Per 100000 to White Population Percentage', 
    x = 'White percentage of total population', 
    y = 'Case rate per 100000',
    color = 'Log 10 of population density'
  )
```

```{r message=FALSE}
df %>% 
  ggplot(aes(x = white_population_pct, y = case_rate_100k, color = poverty_pct)) + 
  geom_point() +
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to White Population Percentage', 
    x = 'White percentage of total population', 
    y = 'Case rate per 100000',
    color = 'Poverty percentage'
  )
```

```{r message=FALSE}
df %>% 
  ggplot(aes(x = poverty_pct , y = case_rate_100k, color = log10(population_density))) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to Poverty Percentage', 
    x = 'Poverty Percentage', 
    y = 'Case rate per 100000',
    color = 'Log 10 of population density'
  )
```

```{r message=FALSE}
df %>% 
  ggplot(aes(unemployed_pct, case_rate_100k, color = log10(population_density))) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to Unemployed Percentage', 
    x = 'Unemployed percentage (2018)', 
    y = 'Case rate per 100000',
    color = 'log 10 of population density'
  )
```

```{r message=FALSE}
df %>% 
  ggplot(aes(senior_pct, case_rate_100k, color = log10(population_density))) + 
  geom_point() +   
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to Seniors Percentage (65+)', 
    x = 'Seniors percentage', 
    y = 'Case rate per 100000',
    color = 'log 10 of population density'
  )
```

### Model 2 Regression
The regression models regarding the demographic variables discussed previously are listed as below.

@@

```{r}
model2 <- lm(case_rate_100k ~ senior_per_100k + poverty_per_100k + unemployed_per_100k, data = df)
coeftest(model2)
```

#### (need to update)
From the results of the regression, none of the coefficients of investigated variables in model 2 are significant (p-value < 0.05). Among these regression coefficient tests,  the p-value regarding the poverty percentage variable is the lowest (0.05755).  In addition, its positive estimated coefficient is consistent with the statement that low-income communities are at high pandemic risk claimed by mainstream media these days.

### Model 2 limits

#### i. IID Sampling 
( Discussed in the common section)

#### ii.Linear Conditional Expectation
Assessing non-linear in higher-dimensional space: look at the predicted vs. residuals of the model.

```{r, echo=FALSE}
df<- df%>% 
  mutate(
    model2_preds = predict(model2), 
    model2_resids = resid(model2)
  ) 
df %>% 
  ggplot(aes(model2_preds, model2_resids)) + 
  geom_point() + 
  stat_smooth()
```

####iii.No Perfect Collinearity
First, check if any variables were dropped by R. No coefficient is missing, which mean there is no perfect collinearity.

@@

In addition, all the variance inflation factors are less than 4, which doesn't indicate the exisitence of collinearity.

@@

####iv.Homoskedastic Errors
10 is Florida, 33 is New York, 35 is North Dakota

@@

@@

####v.Normally Distributed Errors
Check the normality of error distribution based on the histogram of residuals and the qqplot.

```{r, echo=FALSE, warning=FALSE}
plot_one <- df %>% 
  ggplot(aes(x = model2_resids)) + 
  stat_bin(bins = 25)
  geom_histogram()

@@

plot_one / plot_two
```

#### GGPairs

```{r, echo=FALSE}
df %>% 
  select(case_rate_100k, senior_per_100k, poverty_per_100k, unemployed_per_100k) %>% 
  GGally::ggpairs()
```

#### Cook Distance 
10 Florida, 20 Maine, 35 North Dakota

```{r, echo =FALSE}
plot(model2, which = 5)
```

>>>>

df <-df %>%
  mutate(    
    senior_per_100k = 100000*senior_pct,
    poverty_per_100k = 100000*poverty_pct/100, 
    unemployed_per_100k = 100000*unemployed_pct/100,
  )
```

@@

```{r, warning=FALSE, message=FALSE}

ggplot(data = df,
  mapping = aes(x= senior_per_100k)) +
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(10000,21000,1000)) +
  labs(title = 'Histogram of Seniors Distribution', 
  x = 'Number of seniors per 100,000', y = 'Count') + 
  scale_x_continuous(breaks=seq(10000,22000,1000))
```

@@

Based on the work of model 1, two extra state demographic information regarding poverty rate (percentage living under the federal poverty line in 2018) and unemployment rate (percentage of unemployed in 2018) would be included in the analysis of model 2. In addition, both newly-introduced percentage variables would be transformed into rates per 100,000 to align with the case rate variable.

First, the distribution of the poverty rate is examined as below.

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6000,21200,800)) +
  labs(title =  "Histogram of Poverty Rate", 
       x = "Number of living under poverty line per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(6000, 21200, 1600))
```
As shown in the above distribution, there is a fairly normal distribution in the poverty number 
per 100,000 across the states. Additionally, there is no outliers observed within the distribution,
given that the range of rates fall between 7,600 and 19,700. Therefore, there is no further transformation in need for this variable in the following analysis process for both model 2 and model 3.

Next, the distribution of the unemployment rate is examined as below.

```{r}
ggplot(data = df,
  mapping = aes(x= unemployed_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(2000,8000,500)) +
  labs(title =  "Histogram of Unemployment Rate", 
       x = "Number of unemployed per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(2000, 8000, 1000))
```
According to the histogram above, the distribution of unemployed number per 100,000 is of fairly normal distribution. Overall, it is not highly skewed or heavily tailed. Therefore, there is no further transformation in need for this variable in the following analysis process for both model 2 and model 3.

#### Model 2 Regression

Compared to model 1, the relationship investigated here is about variables space with higher dimensions. It would be challenging to observe notable variation from variables distribution plots as below.

```{r message=FALSE}
case_poverty <- df %>% 
  ggplot(aes(poverty_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = ' Association between Covid-19 ase rate and the poverty rate', 
    x = 'Number of living under poverty line per 100,000',
    y = 'Covid-19 care rate per 100,000', 
    color = 'Senior num per 100k'
  )

case_unemployed <- df %>% 
  ggplot(aes(unemployed_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Association between Covid-19 ase rate and the unemployment rate', 
    x = 'Number of unemployed per 100,000', 
    y = 'Covid-19 care rate per 100,000',
    color = 'Senior num per 100k'
  )

case_poverty / case_unemployed
```
However, there is an overall increasing trend of COVID-19 care rate with increasing poverty rate, while there is a inverse relationship between the variables of the COVID-19 case rate and the unemployment rate.

Based on above data exploratory result, the following equation is used to create a regression model
to determine the relationship between the case rate and demographic features including the seniors rate, the rate of poverty and the rate of unemployment.

@@

```{r}
model2 <- lm(case_rate_100k ~ senior_per_100k + poverty_per_100k + unemployed_per_100k, data = df)
```

Accordingt to the assessment of homoskedasticity of model 2, which would be discussed in the following paragraph (Model 2 limits - iv.Homoskedastic Errors), it is plausible to apply the classical standard errors in the t-test of coefficients here.

```{r}
coeftest(model2)
```

Based on the test results of the estimated coefficients in model 2, all three demographic features discussed here are significantly realated to the COVID-19 care rate, especially for the poverty rate (p-value < 0.01).

The coefficient of the senior_per_100k is  -0.230970 and statistically significant, which implies that for every one thousand additional seniors per 100,000, there is a decrease of 231 COVID-19 cases per 100,000 holding other demographic features as constant. As discussed in mode1 1 section, although seniors are susceptible group, less mobility and social interaction with cautious self procetion may play a important role in this pandemic situation.

In addition, the coefficient of the poverty_per_100k (0.231750) indicats that for every one thousand additional people living under poverty line per 100,000, there is a increase of 232 COVID-19 case per 100,000 holding other demographic features as constant. People living under the poverty line have limited accesss to medical resourses. Moreover, most of them could only afford to live in the communites with poor hygiene.  condition.

Finally, the coefficient of the unemployed_per_100k is statiscally significant and negative(-0.486191), which implies that for every one thousand additional unemployed, there is a decrease of 486 COVID-19 case per 100,000 holding other demographic features as constant. Although the unemployment leads to less medical insurace coverage and impact on the incoming, the unempolyed people may have more flexibility in obeying stay-at-home. In addition, there is less expourse probability when no commuting or in-place working is required for the unemployed.

#### Model 2 limits

##### i. IID Sampling 
(Would be discussed in the following section - 3. Limitations of the Models)

##### ii.Linear Conditional Expectation
To assess the linear conditional expectation assumption in higher-dimensional space, the plots of predicted versus residuals of the model 2 is investigated as follow.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
df<- df%>% 
  mutate(
    model2_preds = predict(model2), 
    model2_resids = resid(model2)
  ) 
df %>% 
  ggplot(aes(model2_preds, model2_resids)) + 
  geom_point() + 
  stat_smooth()
```

Overall, the residuals maintain around zero across the predicted range. There is no assumption violation phenomenon could be observed here.

#####iii.No Perfect Collinearity

Since there are multiple covariates included in this model compared to the model 1, there is more concern about the no perfect collinearity assumption here
First, the list of variables coefficients shows that no variables were dropped, which mean there is no perfect collinearity detected by the R function.

@@

In addition, all the variance inflation factors are less than 4 as follow, which doesn't indicate the exisitence of collinearity.

@@

#####iv.Homoskedastic Errors
To evaluate the homoskedastic errors assumption, the square root of the residuals is plotted against the fitted values of the model 2 (scale-location plot). Although, the curve is not perfect flat smooth, there is no obvious variance of the errors either. For reference, the data points marked in the plot represent Florida(10), New York (33) and North Dakota (35).

@@

Additionally, a Breusch-Pagan test is run to check the level of heteroskedasticity. The test result shows it fails to reject the null hypotheis, which means there is no evidence for heteroskedasticity. Overall, the assumption of homoskedastic errors is satisfied for model 2.

@@

#####v.Normally Distributed Errors
To access the assumption of normality of error distribution, both the histogram and the qqplot of the residuals is shown as follow.

```{r, echo=FALSE, warning=FALSE, message=False}
plot_one <- df %>% 
  ggplot(aes(x = model2_resids)) +
  geom_histogram(fill = 'skyblue', color = 'grey30', bins=30) +
  labs(title =  "Histogram of the Model 2 Residuals", 
       x = "Model 2 residuals", y = 'Count')

@@

plot_one / plot_two
```
Limited by the sample size, it is hard to observe a perfect normal distribution from the histogram. However, the qqplot implies no deviation from normality. Overall, the normally distributed errors assumption is satisfied for model 2.

##### Cook Distance

Additionally, cook distance is used to estimate the influence of each single data point.

```{r, echo =FALSE}
plot(model2, which = 5)
```
No obvious outlier could be detected here. For reference, the marked data points are Florida(10), Maine(20), and North Dakota(30).
